### 2015-06-24 Dissertation/abstract.tex
Grammar fixes from Erik

The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
evolutionary optimization without requiring any user specified parameters.
P3's primary innovation is to replace the generational model with a pyramid of
multiple populations that are iteratively created and expanded. In combination
with local search and advanced crossover,
P3 scales to problem difficulty, exploiting
previously learned information before adding more diversity.

@@

Unlike the best comparison algorithms, P3 was able to achieve this quality
\textit{without any
problem specific tuning}. Thus, unlike previous parameter-less methods, P3 does not
sacrifice quality for applicability. Therefore we conclude that
P3 is an efficient, general, parameter-less approach to black-box
optimization that is more effective than existing state-of-the-art techniques.

>>>>

The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
evolutionary optimization without requiring any user-specified parameters.
P3's primary innovation is to replace the generational model with a pyramid of
multiple populations that are iteratively created and expanded. In combination
with local search and advanced crossover,
P3 scales to problem difficulty, exploiting
previously learned information before adding more diversity.

@@

Unlike the best comparison algorithms, P3 was able to achieve this quality
\textit{without any
problem-specific tuning}. Thus, unlike previous parameter-less methods, P3 does not
sacrifice quality for applicability. Therefore we conclude that
P3 is an efficient, general, parameter-less approach to black-box
optimization that is more effective than existing state-of-the-art techniques.

### 2015-06-24 Dissertation/chapter-comparison_algorithms.tex
Grammar fixes from Erik

In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms that
have related features to P3. The Random Restart Hill Climber defined by ~\cite{goldman:2014:p3} was chosen as an efficient form of repeated
local search. As P3 combines this hill climber with crossover,
comparing with local search alone shows the advantages of P3's overall approach.
The $1+(\lambda, \lambda)$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
hBOA are advanced model building search techniques that are effective at learning complex
problem structure, designed to achieve similar goals as P3's linkage learning but using very different
methods. Finally LTGA represents the current state-of-the-art in black-box search and is the origin
of P3's linkage learning and crossover methods.

@@

%%%%%%%%%% HC %%%%%%%%%%%%%%%%
\section{Random Restart Hill Climber}
\label{sec-hill-climber}
Perhaps the simplest black-box search heuristic is stochastic local
search, also know as hill climbing.
This optimization technique focuses on improving a single solution until it reaches a local
optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
order, keeping modifications when fitness is improved, until single bit flips cannot result in
further fitness improvements.

@@

%%%%%%%%%%%% Lambda Lambda %%%%%%%%%%%%%%%%
\section{$1+(\lambda, \lambda)$}
\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
the advantages of performing crossover on
% bwg: Instead of "the problem known as" how about "the One Max" problem?
the problem known as % wfp
One Max. 
% wfp could use a reference for OneMax
% bwg: I've looked, and One Max seems to be so common that no one cites a seminal paper for it.
This comparatively simple
algorithm maintains only a single individual and a self-controlled parameter $\lambda$.

@@

The original formulation was designed specifically for unimodal landscapes and as such were
not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
reset to 1.

@@

Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
relationships that should be preserved.
The process begins by creating the set of sets $unmerged$ that tracks all top level clusters. Initially
$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
until only a single set remains in $unmerged$ which contains all of the genes in the genome.

@@

\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
using only high quality solutions, even during the first generation.

@@

In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
unhelpful, and all donations were evaluated, even those that did not change any genes, then \Call{Cluster-Usage}{}
would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
required to make clustering only \BigO{N} operations per evaluation. However, by skipping some evaluations, its
possible that clustering may be super-linear.

@@

By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
the population size, relying instead on a fixed user specified size.

@@

While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each problem size.
Extended by \cite{goldman:2012:ltga}, this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the 
smallest successful and the % bwg: added "the"
largest unsuccessful sizes. In this way the minimum population size
that meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run that had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

While not currently treated as a parameter, we also performed preliminary tests of
integrating hill climbing into LTGA and hBOA
as this is the procedure used in P3.
To match P3, we applied first improvement
hill climbing to each algorithm's initial population.
We then performed bisection on the modified algorithms for the largest problem sizes
where hBOA without hill climbing was effective. We found that in general both methods
performed worse when combined with hill climbing, in some cases up to an order of
magnitude worse. There were three exceptions: both improved on MAX-SAT and
hBOA improved on Rastrigin. In all cases the inclusion of hill climbing did
not result in either algorithm outperforming P3 in terms of evaluations
required to reach the global optimum. As such, all further experiments
use the unmodified,
published versions.

>>>>

In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms that
have related features to P3. The Random Restart Hill Climber defined by ~\cite{goldman:2014:p3} was chosen as an efficient form of repeated
local search. As P3 combines this hill climber with crossover,
comparing with local search alone shows the advantages of P3's overall approach.
The $(1+(\lambda, \lambda))$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
hBOA are advanced model building search techniques that are effective at learning complex
problem structure, designed to achieve similar goals as P3's linkage learning but using very different
methods. Finally LTGA represents the current state-of-the-art in black-box search and is the origin
of P3's linkage learning and crossover methods.

@@

%%%%%%%%%% HC %%%%%%%%%%%%%%%%
\section{Random Restart Hill Climber}
\label{sec-hill-climber}
Perhaps the simplest black-box search heuristic is stochastic local
search, also know as hill climbing.
This optimization technique focuses on improving a single solution until it reaches a local
optimum. Here we use the first-improvement hill climber defined by \cite{goldman:2014:p3}
and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
order, keeping modifications when fitness is improved, until single bit flips cannot result in
further fitness improvements.

@@

%%%%%%%%%%%% Lambda Lambda %%%%%%%%%%%%%%%%
\section{$(1+(\lambda, \lambda))$}
\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
the advantages of performing crossover on
% bwg: Instead of "the problem known as" how about "the One Max" problem?
the problem known as % wfp
One Max. 
% wfp could use a reference for OneMax
% bwg: I've looked, and One Max seems to be so common that no one cites a seminal paper for it.
This comparatively simple
algorithm maintains only a single individual and a self-controlled parameter $\lambda$.

@@

The original formulation was designed specifically for unimodal landscapes and as such were
not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $(1+(\lambda, \lambda))$
to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
reset to 1.

@@

Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
relationships that should be preserved.
The process begins by creating the set of sets $unmerged$ that tracks all top-level clusters. Initially
$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
until only a single set remains in $unmerged$ which contains all of the genes in the genome.

@@

\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
using only high-quality solutions, even during the first generation.

@@

In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
unhelpful, and all donations were evaluated, even those that did not change any genes, then \Call{Cluster-Usage}{}
would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
required to make clustering only \BigO{N} operations per evaluation. However, by skipping some evaluations, it is
possible that clustering may be super-linear.

@@

By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
the population size, relying instead on a fixed user-specified size.

@@

While four of the six algorithms in our experiments do not require any user-specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each problem size.
Extended by \cite{goldman:2012:ltga}, this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the 
smallest successful and the % bwg: added "the"
largest unsuccessful sizes. In this way the minimum population size
that meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run that had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

While not currently treated as a parameter, we also performed preliminary tests of
integrating hill climbing into LTGA and hBOA
as this is the procedure used in P3.
To match P3, we applied first-improvement
hill climbing to each algorithm's initial population.
We then performed bisection on the modified algorithms for the largest problem sizes
where hBOA without hill climbing was effective. We found that in general both methods
performed worse when combined with hill climbing, in some cases up to an order of
magnitude worse. There were three exceptions: both improved on MAX-SAT and
hBOA improved on Rastrigin. In all cases the inclusion of hill climbing did
not result in either algorithm outperforming P3 in terms of evaluations
required to reach the global optimum. As such, all further experiments
use the unmodified,
published versions.

### 2015-06-24 Dissertation/chapter-conclusions.tex
Grammar fixes from Erik

Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
quality extends to intermediate points during evolution, with P3 generally reaching at least as high
of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
is some evidence that even when compared on wall clock time, P3 performs on par with
the best comparison techniques. All of these achievements are made without any problem specific parameter
tuning, making P3 easier to apply to new domains than its two closets competitors in quality.

P3's quality is due to a number of desirable traits. First, mixing local search with model based
crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques that must
overcompensate with larger population sizes on single instance problems and consider the worst instance
when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the
nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
each fitness level, similar to a generational model using a decreasing population size.

@@

There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
pressing for practitioner acceptance is to apply P3 to real world problems and compare
its results with other black-box or even problem specific heuristics. While parameter-less,
P3 is currently limited to discrete, fixed length genomes evaluated using single objective
fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
While asymptotically linear in problem size, Black-Box P3's modeling techniques and local search methods are likely going
to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
and the inability of the model to capture overlapping linkage may be hindering
search efficiency. Overcoming these limitations by using a new modeling technique may allow
the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
order deception via linkage learning, the iterative improvement method by which crossovers
are made may mislead search on landscapes with higher order deception. This appears
to be especially important on NP-Hard problems, where high quality solutions may be
very distant from the global optimum in representation space.

While empirically effective, we believe its possible to derive theoretical foundations
for Gray-Box P3. Specifically, this may be possible by considering the nature of crossover
in terms similar to those use for finding $r$-bit local optima with HBHC. Consider
that for any two parents the non-linear effects of crossover are constrained by
where they containing different bit values. As a result, subsets of their differences
can become linearly separable, allowing for each to be considered independently of all others.
Furthermore, these conclusions about effective crossover in gray-box can likely be translated
into a black-box setting to help explain the effectiveness of Black-Box P3.

However, even without
these improvements our results show P3 is highly efficient at finding
global optima on both black-box and gray-box problems without any problem specific tuning.

>>>>

Across a large number of problems and problem sizes P3 required fewer evaluations to reach the global optimum
than optimally tuned state-of-the-art competitors. On single-instance problems P3's improvement was by a
constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
quality extends to intermediate points during evolution, with P3 generally reaching at least as high
a fitness as the competitive techniques when using the same number of evaluations. While P3 does require
modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
is some evidence that even when compared on wall clock time, P3 performs on par with
the best comparison techniques. All of these achievements are made without any problem-specific parameter
tuning, making P3 easier to apply to new domains than its two closest competitors in quality.

P3's quality is due to a number of desirable traits. First, mixing local search with model based
crossover lets search focus on properly mixing high-quality solutions. Second, by adding diversity only
as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques that must
overcompensate with larger population sizes on single-instance problems and consider the worst instance
when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
to reach high-quality intermediate fitnesses quickly without prematurely converging. Fourth, the
nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
each fitness level, similar to a generational model using a decreasing population size.

@@

There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
pressing for practitioner acceptance is to apply P3 to real-world problems and compare
its results with other black-box or even problem-specific heuristics. While parameter-less,
P3 is currently limited to discrete, fixed-length genomes evaluated using single-objective
fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
While asymptotically linear in problem size, Black-Box P3's modeling techniques and local search methods are likely going
to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
and the inability of the model to capture overlapping linkage may be hindering
search efficiency. Overcoming these limitations by using a new modeling technique may allow
the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low-order
deception via linkage learning, the iterative improvement method by which crossovers
are made may mislead search on landscapes with higher-order deception. This appears
to be especially important on NP-Hard problems, where high-quality solutions may be
very distant from the global optimum in representation space.

While empirically effective, we believe it is possible to derive theoretical foundations
for Gray-Box P3. Specifically, this may be possible by considering the nature of crossover
in terms similar to those used for finding $r$-bit local optima with HBHC. Consider
that for any two parents the non-linear effects of crossover are constrained by
where they contain different bit values. As a result, subsets of their differences
can become linearly separable, allowing for each to be considered independently of all others.
Furthermore, these conclusions about effective crossover in gray-box can likely be translated
into a black-box setting to help explain the effectiveness of Black-Box P3.

However, even without
these improvements our results show P3 is highly efficient at finding
global optima on both black-box and gray-box problems without any problem-specific tuning.

### 2015-06-24 Dissertation/chapter-gray-box-p3.tex
Grammar fixes from Erik

In effect the clusters are sampling moves that the HBHC
would make if $r \ge l$. For any solution in the search space that is not globally optimal, there
must exist some move that will improve its quality. However, this move may
be arbitrarily large and it is intractable to test all possible moves of even moderate size.
By sampling from all possible large moves, we can maintain tractability while gaining
a potentially non-zero probability of improvement.
As clusters are used to donate values between solutions, these
moves are always in the direction of previously found high quality solutions.
This assumes that the density of high quality solutions is higher than average between good solutions.

Another efficiency gain is that each time a
donation is made, only the effected part of the solution's fitness needs to be
recalculated. As a result, the number of subfunction evaluations required
to determine the change in fitness is only \BigO{l}. This also
allows for Gray-Box P3 to efficiently reapply hill climbing after each
donation as only effected moves need to be rechecked. As a result, a donation
and its resulting modifications from hill climbing are kept only if the
new local optima is at least as fit as the solution before the donation occurred.
All combined a single donation plus returning to a local optimum requires \BigO{l + I} time,
while just the donation in Black-Box P3 requires \BigO{N}.

>>>>

In effect the clusters are sampling moves that the HBHC
would make if $r \ge l$. For any solution in the search space that is not globally optimal, there
must exist some move that will improve its quality. However, this move may
be arbitrarily large and it is intractable to test all possible moves of even moderate size.
By sampling from all possible large moves, we can maintain tractability while gaining
a potentially non-zero probability of improvement.
As clusters are used to donate values between solutions, these
moves are always in the direction of previously found high-quality solutions.
This assumes that the density of high-quality solutions is higher than average between good solutions.

Another efficiency gain is that each time a
donation is made, only the affected part of the solution's fitness needs to be
recalculated. As a result, the number of subfunction evaluations required
to determine the change in fitness is only \BigO{l}. This also
allows for Gray-Box P3 to efficiently reapply hill climbing after each
donation as only affected moves need to be rechecked. As a result, a donation
and its resulting modifications from hill climbing are kept only if the
new local optimum is at least as fit as the solution before the donation occurred.
All combined, a single donation plus returning to a local optimum requires \BigO{l + I} time,
while just the donation in Black-Box P3 requires \BigO{N}.

### 2015-06-24 Dissertation/chapter-problems.tex
Grammar fixes from Erik

\section{Single Instance Problems}
Understanding how a stochastic search algorithm will behave on arbitrary and complex
search landscapes can be exceedingly difficult. Therefore a common practice for
algorithm understanding is to perform search on well defined, well understood
landscapes. To be of interest these landscapes need to represent interesting
and important aspects of real world problems.

@@

Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
and only If (HIFF) problem~\cite{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
leaf and each internal node is the subset of genes contained in its children. If all genes represented
in a node of the tree are set to the same value, they score equal to the size of the set. In this way
small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.

@@

%%%%%%%%%%%%%%%%%% Randomly Generated %%%%%%%%%%%%%%%%%%%%%%
\section{Randomly Generated Problem Classes}
While well defined landscapes can provide specific insights into how an algorithm works,
their static nature can be misleading. Specifically, algorithm quality might
be so fragile that it is only effective at searching well behaved landscapes. A more realistic
test of an algorithm's black-box effectiveness is to work with randomly generated instances
drawn from a problem class. When tested over a sufficiently large sample it is then possible
to draw more general conclusions about the algorithms effectiveness. The challenge with these
landscapes is determining the global optimum to gauge if an algorithm was successful.

@@

\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
\begin{equation}
\sum_{e_{ij} \in E} x_ie_{ij}x_j
  \label{eq-ising}
\end{equation}
Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
to either -1 or 1, and vertex values must be -1 or 1.

As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
Related to the more common 3-SAT problem, a MAX-SAT instance is
defined by a set of 
% bwg: Why -- (two hypens) and not - (one hypen)?
three--term
clauses.
Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.

>>>>

\section{Single Instance Problems}
Understanding how a stochastic search algorithm will behave on arbitrary and complex
search landscapes can be exceedingly difficult. Therefore a common practice for
algorithm understanding is to perform search on well defined, well understood
landscapes. To be of interest these landscapes need to represent interesting
and important aspects of real-world problems.

@@

Another challenging aspect of landscapes can be higher-order relationships. The Hierarchical If
and only If (HIFF) problem~\cite{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
leaf and each internal node is the subset of genes contained in its children. If all genes represented
in a node of the tree are set to the same value, they score equal to the size of the set. In this way
small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
all 1s or all 0s, meaning that to solve higher-order subproblems it is necessary to perform crossovers
that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.

@@

%%%%%%%%%%%%%%%%%% Randomly Generated %%%%%%%%%%%%%%%%%%%%%%
\section{Randomly Generated Problem Classes}
While well defined landscapes can provide specific insights into how an algorithm works,
their static nature can be misleading. Specifically, algorithm quality might
be so fragile that it is only effective at searching well behaved landscapes. A more realistic
test of an algorithm's black-box effectiveness is to work with randomly generated instances
drawn from a problem class. When tested over a sufficiently large sample it is then possible
to draw more general conclusions about an algorithm's effectiveness. The challenge with these
landscapes is determining the global optimum to gauge if an algorithm was successful.

@@

\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
\begin{equation}
\sum_{e_{ij} \in E} x_ie_{ij}x_j
  \label{eq-ising}
\end{equation}
Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
In this subset the graph is restricted to be a two-dimensional torus, edge weights are randomly set
to either -1 or 1, and vertex values must be -1 or 1.

As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
Related to the more common 3-SAT problem, a MAX-SAT instance is
defined by a set of 
% bwg: Why -- (two hypens) and not - (one hypen)?
three--term
clauses.
Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
problem is challenging we chose a clause-to-variable ratio of $4.27$~\citep{selman:1996:maxsat}.
